---
title: "DemoKin: An R package to compute kinship networks in stable and non-stable populations"
date: "Formal Demography Working Group - Feb 25 2022"
author: |
   | Ivan Williams (Universidad de Buenos Aires)
   | Diego Alburez-Gutierrez* (Max Planck Institute for Demographic Research)
header-includes:
  - \usepackage{amsmath}
bibliography: kinship.bib
output:
  bookdown::html_document2:
    keep_md: true
    number_sections: true
    code_folding: show
    toc: true
    toc_depth: 2
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("functions.R")

# Install packages if needed
library(DemoKin)
library(tidyverse)
library(knitr)
library(fields)
```

> **Get the `Rmd` version of this file: https://github.com/alburezg/kinship_formal_demo.**

# The DemoKin package

`DemoKin` uses matrix demographic methods to compute expected (average) kin counts from demographic rates under a range of scenarios and assumptions. 
The package is an R-language implementation of Caswell [-@caswell_formal_2019;-@caswell_formal_2021] and draws on previous theoretical development by Goodman, Keyfitz and Pullum [-@goodman_family_1974]. 

## Installation

You can install the development version from GitHub with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IvanWilli/DemoKin")
```

## The function `kins()`

`DemoKin::kins()` currently does most of the heavy lifting in terms of implementing matrix kinship models. 
This is what it looks like in action, in this case assuming demographic stability:

```{r}
library(DemoKin)

stable <- 
  kins(
    U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    , ego_year = 1950
    , selected_kins = NULL
    , living = TRUE
    )
```

### Arguments

- **U** matrix; survival ratio by age from a life table 
- **f** matrix; age specific fertility rate by age (simple ages). 
- **N** matrix; only needed for non-stable computation 
- **stable** logical string 
- **ego_year** string; period years of interest
- **ego_cohort** string; birth cohorts of interest
- **pi** matrix; distribution of ages of mothers (see Caswell [@-caswell_formal_2019])
- **birth_female** numeric; proportion of births that are female
- **selected_kins** string; relatives to compute. If `NULL`, return values for all relatives
- **living** logical; whether to compute values for living or deceased relatives 

Note that `DemoKin` only requires period demographic rate data as input!

### Details

Relatives for the `selected_kins` argument are identified by a unique code:

```{r, fig.height=6, fig.width=8, echo=FALSE}

relatives <- c(
  "coa" = "Cousins from older aunt"
  , "cya" = "Cousins from younger aunt"
  , "d" = "Daughter"
  , "gd" = "Grand-daughter"
  , "ggd" = "Great-grand-daughter"
  , "ggm" = "Great-grandmother"
  , "gm" = "Grandmother"
  , "m" = "Mother"
  , "nos" = "Nieces from older sister"
  , "nys" = "Nieces from younger sister"
  , "oa" = "Aunt older than mother"
  , "ya" = "Aunt younger than mother"
  , "os" = "Older sister"
  , "ys" = "Younger sister"
  )

data.frame(Code = names(relatives), Relative = relatives, row.names = NULL) %>% kable()
```

## Built-in data

The `DemoKin` package includes data from Sweden as an example. 
The data comes from the [Human Mortality Database](https://www.mortality.org/) and [Human Fertility Database](https://www.humanfertility.org/). 
These datasets were loaded using the`DemoKin::get_HMDHFD` function.
To list the data:

```{r, eval = FALSE}
data(package="DemoKin")
```

```{r, echo=FALSE}
data(package="DemoKin")$results[4:6,3:4] %>% 
  data.frame() %>% 
  kable()
```

The in-built data sets are:

- **swe_surv** matrix; survival ratio by age (DemoKin's *U* argument)

This is what the data looks like:

```{r}
swe_surv[1:4, 1:4]
```

And plotted over time and age:

```{r}
library(fields)

image.plot(
  x = as.numeric(colnames(swe_surv))
  , y = 0:nrow(swe_surv)
  , z = t(as.matrix(swe_surv))
  , xlab = "Year"
  , ylab = "Survival ratio (U)"
  )
```

- **swe_asfr** matrix; age specific fertility rate (DemoKin's *f* argument)

This is what the data looks like:

```{r}
swe_asfr[1:4, 1:4]
```

And plotted over time and age:

```{r}
image.plot(
  x = as.numeric(colnames(swe_asfr))
  , y = 0:nrow(swe_asfr)
  , z = t(as.matrix(swe_asfr))
  , xlab = "Year"
  , ylab = "Age-specific fertility (f)"
  )
```

- **swe_pop** matrix; population by age (Demokin's *N* argument)

This is what the data looks like:

```{r}
swe_pop[1:4, 1:4]
```

And plotted over time and age:

```{r}
image.plot(
  x = as.numeric(colnames(swe_pop))
  , y = 0:nrow(swe_pop)
  , z = t(as.matrix(swe_pop))
  , xlab = "Year"
  , ylab = "Population totals by age (N)"
  )
```

## Value

`DemoKin::kins()` returns a list with two elements.

- **kins_by_age_ego** is a data frame of ...

```{r}
head(stable$kins_by_age_ego)
```

- **kins** is a data frame of ...

```{r}
head(stable$kins)
```

We'll see what these means in the examples below.

# Example 1: kin counts in stable populations

Following Caswell [-@caswell_formal_2019], we assume a female closed population in which everyone experiences the Swedish 1950 mortality and fertility rates at each age throughout their life.
We then ask:

> How can we characterize the kinship network of an average member of the population (call her 'Ego')?

For this exercise, we'll use the Swedish data pre-loaded with `DemoKin`.

```{r}
library(DemoKin)

system.time(
  stable <- 
  kins(
    ego_year = 1950
    , U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    )
)
```

## 'Keyfitz' kinship diagram

We can visualize the implied kin counts for an Ego aged 35 yo in a stable population using a network or 'Keyfitz' kinship diagram [@Keyfitz2005] using the `plot_diagram` function:

```{r, fig.height=10, fig.width=12}
stable$kins_by_age_ego %>% 
  filter(age_ego == 35) %>% 
  select(kin, total) %>% 
  plot_diagram()
```


## Expected kin counts for an Ego surviving to each age

before showing the results, we define a simple function to identify relatives based on their respective codes.
The kinship codes (e.g., "gm", "ggm") are useful for filtering the data but confusing for visualisation. 

```{r}
rename_kin <- function(df){
  df$kin <- relatives[df$kin]
  df
}
```

Now, let's visualize how the expected number of daughters, siblings, cousins, etc., changes over the lifecourse of Ego.

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total))  +
  geom_vline(xintercept = 35, color=2)+
  theme_bw() +
  labs(x = "Ego's age") +
  facet_wrap(~kin)
```

It's a bit too much information in one graph. Let's pick it appart by considering different kin types in isolation. 

- **Living ancestors**

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("m", "gm", "ggm")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving kin")
```

- **Living descendants**

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("d", "gd", "ggd")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving kin")
```

- **Other kin**

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("os", "ys", "cya", "coa", "ya", "oa")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving kin")
```

- **Family size**: How does overall family size (and family composition) vary over life for an average woman who survives to each age?

```{r}
stable$kins_by_age_ego %>%
  group_by(age_ego) %>% 
  summarise(total = sum(total)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(age_ego, total), size = 1)  +
  coord_cartesian(ylim = c(0, 6)) + 
  theme_bw() +
  labs(x = "Ego's age", y = "Number of living female relatives")
```

We can decompose this by relative type:

```{r, fig.height=6, fig.width=8}

# Combine kin types irrespective of whether they come from older
# or younger sibling lines
consolidate <- c("c", "c", "d", "gd", "ggd", "ggm", "gm", "m", "n", "n", "a", "a", "s", "s")
names(consolidate) <- c("coa", "cya", "d", "gd", "ggd", "ggm", "gm", "m", "nos", "nys", "oa", "ya", "os", "ys")

# Rename kin types from codes to actual words
relatives_small <- c("Cousins", "Daughter", "Grand-daughter", "Great-grand-daughter", "Great-grandmother", "Grandmother", "Mother", "Nieces", "Aunt", "Sister")
names(relatives_small) <-  unique(consolidate)

stable$kins_by_age_ego %>%
  select(age_ego, kin, total) %>% 
  # Consolidate kin types
  mutate(kin = consolidate[kin]) %>% 
  group_by(age_ego, kin) %>% 
  summarise(total = sum(total)) %>% 
  ungroup() %>% 
  # Rename by hand
  mutate(kin = relatives_small[kin]) %>% 
  ggplot() +
  geom_area(aes(x = age_ego, y = total, fill = kin)) +
  labs(x = "Ego's age", y = "Number of living female relatives") +
  coord_cartesian(ylim = c(0, 6)) + 
  theme_bw() +
  theme(legend.position = "bottom")
```

## Deceased kin

We have focused on living kin, but what about relatives who have died? 
We can get the cumulative number of kin deaths experienced by an average Ego surviving to a given age by setting the parameter `living = FALSE`.
This creates a new element `kins_death_by_age_ego` in the output value of `kins()`:

```{r, fig.height=6, fig.width=8}
stable_death <- 
  kins(
    ego_year = 1950
    , U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    , living = FALSE
    )
```

For this example, we combine all kin types to show the cumulative burden of kin death for an average member of the population surviving to each age:

```{r}
stable_death$kins_alive_death %>%
  filter(alive =="no") %>%
  group_by(age_ego) %>% 
  summarise(total_cum = sum(total_cum)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(age_ego, total_cum), size = 1)  +
  labs(x = "Ego's age", y = "Number of kin deaths experienced (cumulative)") +
  theme_bw()
```

We can decompose this by relative type:

```{r}
stable_death$kins_alive_death %>%
  filter(alive =="no") %>%
  # Consolidate kin types
  mutate(kin = consolidate[kin]) %>% 
  group_by(age_ego, kin) %>% 
  summarise(total = sum(total_cum)) %>% 
  ungroup() %>% 
  # Rename by hand
  mutate(kin = relatives_small[kin]) %>% 
  ggplot() +
  geom_area(aes(x = age_ego, y = total, fill = kin)) +
  labs(x = "Ego's age", y = "Number of kin deaths experienced (cumulative)") +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Age distribution of relatives

How old are Ego's relatives? Using the `kin` data frame, we can visualize the age distribution of Ego's relatives throughout Ego's life. 
As an example, let's pick three points of Ego's life: when she's born (age=0) at the end of her reproductive life (age=50) and when she retires (age=65).

```{r, fig.height=6, fig.width=8}
stable[["kins"]] %>%
  filter(age_ego %in% c(0, 50, 65)) %>% 
  filter(kin %in% c("m", "d", "os", "ys")) %>%
  mutate(age_ego = as.character(age_ego)) %>% 
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_kin, count, colour = age_ego), size = 1) +
  scale_color_discrete("Ego's age") +
  labs(x = "Age of Ego's kin", y = "Age distribution") +
  theme_bw() +
  facet_wrap(~kin)
```

The output of the `DemoKin::kins()` function can also be used to easily determine the mean age Ego's relatives by kin type. 
For simplicity, let's focus on an Ego aged 35 yo and get the mean age (and standard deviation) of her relatives in our stable population. 

```{r}
ages_df <- 
  stable$kins_by_age_ego %>% 
  filter(age_ego == 35) %>% 
  select(kin, mean_age, sd_age)

ma <- 
  ages_df %>% 
  filter(kin=="m") %>% 
  pull(mean_age) %>% 
  round(1)

sda <- 
  ages_df %>% 
  filter(kin=="m") %>% 
  pull(sd_age) %>% 
  round(1)

```

We can say that the mothers of Ego would be, on average, `r ma` years old, with a standard deviation of `r sda` years. 

# Example 2: population with changing rates

The real population is Sweden is not stable: individuals in it experience changing mortality and fertility rates throughout their life.
Kinship structures in populations with changing rates can be computed following Caswell [-@caswell_formal_2021].

All the outputs that we showed for stable populations in the previous section are also available for non-stable populations (e.g., kin counts, deceased kin, and age distributions). 
In this section we'll focus on outputs that are specific to time-varying kinship structures. 
In particular, we'll show period, cohort, and age results for Sweden (Figure 5 from Caswell [-@caswell_formal_2021]). 

Note that, in order to arrive a this results, we use a different set of input rates, which are also pre-loaded with DemoKin. 

```{r}
data(package="DemoKin")
```

```{r, echo=FALSE}
data(package="DemoKin")$results[1:3,3:4] %>% 
  data.frame() %>% 
  kable()
```

## Cohort perspective

For a cohort perspective, we run the `kins()` function with the `stable = FALSE` parameter and with a vector of `ego_year` values:
  
```{r, out.width = '100%'}

system.time(
  swe_coh <- 
    kins(
      U = U
      , f = f
      , pi = pi
      , stable = F
      , birth_female = 1
      # Note that we use the 'ego_cohort' parametersa as input
      , ego_cohort = c(1891,1911,1931,1951,1971,2011,2041)
      # We're only interested in certain kin ties
      , selected_kins = c("d","gd","m","gm")
    )
)
```

Show, in a Keyfitz diagram, the expected kin count for an average woman aged 35 and born in 1951 Sweden in a non-stable population:
  
```{r, fig.height=10, fig.width=12}
swe_coh$kins_by_age_ego %>% 
  filter(cohort == 1951, age_ego == 35) %>% 
  select(kin, total) %>% 
  plot_diagram()
```

Finally, we show the expected number of ascendants and descendants by Ego's birth cohort in any given year:

```{r}
swe_coh$kins_by_age_ego %>%
  rename_kin() %>% 
  mutate(cohort = factor(cohort)) %>% 
  ggplot(aes(year,total,color=cohort)) +
  scale_y_continuous(labels = seq(0,3,.2),breaks = seq(0,3,.2))+
  geom_line(size=1)+
  geom_vline(xintercept = 2019, linetype=2)+
  facet_wrap(~kin,scales = "free")+
  labs(x = "Year", "Kin count") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

## Period perspective

For a **period perspective**, we re-run the `kins()` function, this time with a vector of `ego_year` values:

```{r, out.width = '100%'}
system.time(
swe_period <- 
  kins(
    U = U
    , f = f
    , pi = pi
    , stable = F
    , birth_female = 1
    # Note that we use the 'ego_year' parametersa as input
    , ego_year = c(1891,1921,1951,2010,2050,2080,2120)
    # We're only interested in certain kin ties
, selected_kins = c("d","gd","m","gm","os","ys")
)
)

```

Now, we plot the expected values by Ego's age at different time-points:

```{r}
swe_period$kins_by_age_ego %>%
  rename_kin() %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(age_ego,total,color=year)) +
  geom_line(size=1)+
  facet_wrap(~kin, scales = "free")+
  labs(x = "Ego's age", "Kin count") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

# In the pipeline

1. Hex logo
1. Implement multi-stage [@caswell_formal_2020] and two-sex models [@caswell2021formal_two-sex]
1. Improve Keyfitz kinship diagrams
1. Create a *Human Kinship Database*?

<!-- 1. Functions to transform common demographic data sources (HMD, HFD, UNWPP) to DemoKin input -->
  
# Get involved!
  
`DemoKin` is giving its first steps.
Please contact us via email, or create an issue or submit a pull request on GitHub (https://github.com/IvanWilli/DemoKin).
You can also get in touch directly: 
  
- Ivan: 
    + ivanwilliams1985[at]gmail.com
    + https://twitter.com/ivanchowilliams
- Diego
    + alburezgutierrez[at]demogr.mpg.de
    + https://twitter.com/d_alburez
    + www.alburez.me

# Citation

Williams, Iván, Alburez-Gutierrez, Diego and Xi Song. (2021) "DemoKin: An R package to compute kinship networks in stable and non-stable populations." URL: <https://github.com/IvanWilli/DemoKin>.

# Aknowledgments

Kin count estimation has a long history in mathematical demography, starting with the work of [@lotka1931orphanhood] on modelling orphanhood in theoretical populations across demographic regimes.
To the best of our knowledge, Brass [-@brass_derivation_1953] proposed the first kinship equation to estimate child survival over maternal age. 
Goodman et. al [-@goodman_family_1974] generalized this approach to sisters, granddaughters, cousins, etc., in stable and non-stable populations.
The so-called Goodman-Keyfitz-Pullum Kinship Equations were popularized by Keyfitz and Caswell [-@Keyfitz2005] and Bongaarts [-@bongaarts_projection_1987] used a similar approach to estimate descendants in his 'Family Status Model'.
More recently, Caswell [-@caswell_formal_2019;-@caswell_formal_2021] introduce the matrix approach that `DemoKin` is based on.

# References

# Appendix: Obtaining U matrix from a life table

Here, we show how to estimate the argument **U** from `DemoKin::kins()` (survival ratio by age from a life table) from a standard life table.
The code is part of the function `DemoKin::get_HMDHFD()`.
For this example, we use Swedish period life tables from the Human Mortality Database. 

```{r}
library(tidyverse)

lt <- read.table("fltper_1x10.txt", header = T, skip = 2) 

lt$Age[lt$Age == "110+"] <- 110
lt$Age <- as.numeric(lt$Age)

age = 0:100
ages = length(age)
w = last(age)

# survival probabilities
L <- lt %>%
  filter(Age<=w) %>%
  mutate(Lx = ifelse(Age==w, Tx, Lx)) %>%
  select(Year, Age, Lx) %>%
  spread(Year, Lx) %>%
  select(-Age)

U <- rbind(L[c(-1,-101),]/L[-c(100:101),],
           L[101,]/(L[100,]+L[101,]),
           L[101,]/(L[100,]+L[101,]))
   
rownames(U) <- age
colnames(U) <- unique(lt$Year)
U[is.na(U)] <- 0

U[1:4, 1:4]

```