---
title: "DemoKin: An R package to compute kinship networks in stable and non-stable populations"
date: "Formal Demography Working Group - Feb 25 2022"
author: |
   | Ivan Williams (Universidad de Buenos Aires)
   | Diego Alburez-Gutierrez* (Max Planck Institute for Demographic Research)
header-includes:
  - \usepackage{amsmath}
bibliography: kinship.bib
output:
  bookdown::html_document2:
    keep_md: true
    number_sections: true
    code_folding: show
    toc: true
    toc_depth: 2
    toc_float: 
      collapsed: false
---

<style type="text/css">
  body{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("functions.R")

# Install packages if needed
library(DemoKin)
library(tidyverse)
library(knitr)
library(fields)
```

> **Get the `Rmd` version of this file: https://github.com/alburezg/kinship_formal_demo.**

# The DemoKin package

`DemoKin` uses matrix demographic methods to compute expected (average) kin counts from demographic rates under a range of scenarios and assumptions. 
The package is an R-language implementation of Caswell [-@caswell_formal_2019;-@caswell_formal_2021] and draws on previous theoretical development by Goodman, Keyfitz and Pullum [-@goodman_family_1974]. 

## Installation

You can install the development version from GitHub with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github("IvanWilli/DemoKin")
```

## The function `kins()`

`DemoKin::kins()` currently does most of the heavy lifting in terms of implementing matrix kinship models. 
This is what it looks like in action, in this case assuming demographic stability:

```{r}
library(DemoKin)

stable <- 
  kins(
    U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    , ego_year = 1950
    , selected_kins = NULL
    , living = TRUE
    )
```

### Arguments

- **U** matrix; survival ratio by age from a life table 
- **f** matrix; age specific fertility rate by age (simple ages). 
- **N** matrix; only needed for non-stable computation 
- **stable** logical string 
- **ego_year** string; period years of interest
- **ego_cohort** string; birth cohorts of interest
- **pi** matrix; distribution of ages of mothers (see Caswell [@-caswell_formal_2019])
- **birth_female** numeric; proportion of births that are female
- **selected_kins** string; relatives to compute. If `NULL`, return values for all relatives
- **living** logical; whether to compute values for living or deceased relatives 

Note that `DemoKin` only requires period demographic rate data as input!

### Details

Relatives for the `selected_kins` argument are identified by a unique code:

```{r, fig.height=6, fig.width=8, echo=FALSE}

relatives <- c(
  "coa" = "Cousins from older aunt"
  , "cya" = "Cousins from younger aunt"
  , "d" = "Daughter"
  , "gd" = "Grand-daughter"
  , "ggd" = "Great-grand-daughter"
  , "ggm" = "Great-grandmother"
  , "gm" = "Grandmother"
  , "m" = "Mother"
  , "nos" = "Nieces from older sister"
  , "nys" = "Nieces from younger sister"
  , "oa" = "Aunt older than mother"
  , "ya" = "Aunt younger than mother"
  , "os" = "Older sister"
  , "ys" = "Younger sister"
  )

data.frame(Code = names(relatives), Relative = relatives, row.names = NULL) %>% kable()

```

Let's write a simple function to identify relatives based on their respective codes:

```{r}
rename_kin <- function(df){
  df$kin <- relatives[df$kin]
  df
}
```

## Built-in data

The `DemoKin` package includes data from Sweden as an example. 
The data comes from the [Human Mortality Database](https://www.mortality.org/) and [Human Fertility Database](https://www.humanfertility.org/). 
These datasets were loaded using the`DemoKin::get_HMDHFD` function.
To list the data:

```{r, eval = FALSE}
data(package="DemoKin")
```

```{r, echo=FALSE}
data(package="DemoKin")$results[4:6,3:4] %>% 
  data.frame() %>% 
  kable()
```

The in-built data sets are:

- **swe_surv** matrix; survival ratio by age (DemoKin's *U* argument)

This is what the data looks like:

```{r}
swe_surv[1:4, 1:4]
```

And plotted over time and age:

```{r}
library(fields)

image.plot(
  x = as.numeric(colnames(swe_surv))
  , y = 0:nrow(swe_surv)
  , z = t(as.matrix(swe_surv))
  , xlab = "Year"
  , ylab = "Survival ratio (U)"
  )
```

- **swe_asfr** matrix; age specific fertility rate (DemoKin's *f* argument)

This is what the data looks like:

```{r}
swe_asfr[1:4, 1:4]
```

And plotted over time and age:

```{r}
image.plot(
  x = as.numeric(colnames(swe_asfr))
  , y = 0:nrow(swe_asfr)
  , z = t(as.matrix(swe_asfr))
  , xlab = "Year"
  , ylab = "Age-specific fertility (f)"
  )
```

- **swe_pop** matrix; population by age (Demokin's *N* argument)

This is what the data looks like:

```{r}
swe_pop[1:4, 1:4]
```

And plotted over time and age:

```{r}
image.plot(
  x = as.numeric(colnames(swe_pop))
  , y = 0:nrow(swe_pop)
  , z = t(as.matrix(swe_pop))
  , xlab = "Year"
  , ylab = "Population totals by age (N)"
  )
```

## Value

`DemoKin::kins()` returns a list with two elements.

- **kins_by_age_ego** is a data frame of ...

```{r}
head(stable$kins_by_age_ego)
```

- **kins** is a data frame of ...

```{r}
head(stable$kins)
```

We'll see what these means in the examples below.

# Example 1: kin counts in stable populations

Following Caswell [-@caswell_formal_2019], we assume a female closed population in which everyone experiences the Swedish 1950 mortality and fertility rates at each age throughout their life.
We then ask:

> How can we characterize the kinship network of an average member of the population (call her 'Ego')?

For this exercise, we'll use the Swedish data pre-loaded with `DemoKin`.

```{r}
library(DemoKin)

system.time(
  stable <- 
  kins(
    ego_year = 1950
    , U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    )
)
```

## 'Keyfitz' kinship diagram

We can visualize the implied kin counts for an Ego aged 35 yo in a stable population using a network or 'Keyfitz' kinship diagram [@Keyfitz2005] using the `plot_diagram` function:

```{r, fig.height=10, fig.width=12}
stable$kins_by_age_ego %>% 
  filter(age_ego == 35) %>% 
  select(kin, total) %>% 
  plot_diagram()
```

## Expected kin counts for an Ego surviving to each age

Let's visualize how the expected number of daughters, siblings, cousins, etc., changes over the lifecourse of Ego.

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total))  +
  geom_vline(xintercept = 35, color=2)+
  theme_bw() +
  labs(x = "Ego's age") +
  facet_wrap(~kin)
```

It's a bit too much information. 
Let's consider different kin types in isolation. 

We start by looking at **living ancestors**.

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("m", "gm", "ggm")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving") 
```

Next on, we see **living descendants**.

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("d", "gd", "ggd")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving")
```

We also have kin counts for **horizontal kin**.

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  filter(kin %in% c("os", "ys", "cya", "coa")) %>%
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_ego, total, colour = kin), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Expected number of surviving")
```

How does **overall family size** vary over life for an average woman who survives to each age?

```{r, fig.height=6, fig.width=8}
stable$kins_by_age_ego %>%
  group_by(age_ego) %>% 
  summarise(total = sum(total)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(age_ego, total), size = 1)  +
  theme_bw() +
  labs(x = "Ego's age", y = "Average family size (women only)")
```

## Age distribution of relatives

How old are Ego's relatives? Using the `kin` data frame, we can visualize the age distribution of Ego's relatives throughout Ego's life. 
As an example, let's pick three points of Ego's life: when she's born (age=0) at the end of her reproductive life (age=50) and when she retires (age=65).

```{r, fig.height=6, fig.width=8}
stable[["kins"]] %>%
  filter(age_ego %in% c(0, 50, 65)) %>% 
  filter(kin %in% c("m", "d", "os", "ys")) %>%
  mutate(age_ego = as.character(age_ego)) %>% 
  rename_kin() %>% 
  ggplot() +
  geom_line(aes(age_kin, count, colour = age_ego), size = 1) +
  scale_color_discrete("Ego's age") +
  labs(x = "Age of Ego's kin", y = "Age distribution") +
  theme_bw() +
  facet_wrap(~kin)
```

## Deceased kin

We have focused on living kin, but what about relatives who have died? 
We can get the cumulative number of kin deaths experienced by an average Ego surviving to a given age by setting the parameter `living = FALSE`.
This creates a new element `kins_death_by_age_ego` in the output value of `kins()`:

```{r, fig.height=6, fig.width=8}
stable_death <- 
  kins(
    ego_year = 1950
    , U = swe_surv
    , f = swe_asfr
    , stable = TRUE
    , living = FALSE
    )
```

For this example, we combine all kin types to show the cumulative burden of kin death for an average member of the population surviving to each age:

```{r}
stable_death$kins_alive_death %>%
  filter(alive =="no") %>%
  group_by(age_ego) %>% 
  summarise(total_cum = sum(total_cum)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(aes(age_ego, total_cum), size = 1)  +
  labs(x = "Ego's age", y = "Number of kin deaths experienced (cumulative)") +
  theme_bw()
```

# Example 2: population with changing rates

In the real world, the population is Sweden is not stable.
Individuals in it experience changing mortality and fertility rates throughout their life.
Kinship structures in populations with changing rates can be computed following Caswell [-@caswell_formal_2021].

All the outputs that we showed for stable populations are also available for non-stable populations. 
In this section we'll focus on outputs that are specific to time-varying kinship structures. 
In particular, we'll reproduce Figure 5 from Caswell [-@caswell_formal_2021] to show period, cohort, and age results in Sweden. 

Note that, in order to do this, we use a different set of input rates, which are also pre-loaded with DemoKin. 

```{r}
data(package="DemoKin")
```

```{r, echo=FALSE}
data(package="DemoKin")$results[1:3,3:4] %>% 
  data.frame() %>% 
  kable()
```

## Cohort perspective

For a cohort perspective, we run the `kins()` function with the `stable = FALSE` parameter and with a vector of `ego_year` values:
  
```{r, out.width = '100%'}

system.time(
  swe_coh <- 
    kins(
      U = U
      , f = f
      , pi = pi
      , stable = F
      , birth_female = 1
      # Note that we use the 'ego_cohort' parametersa as input
      , ego_cohort = c(1891,1911,1931,1951,1971,2011,2041)
      # We're only interested in certain kin ties
      , selected_kins = c("d","gd","ggd","m","gm","ggm")
    )
)
```

Show, in a Keyfitz diagram, the expected kin count for an average woman aged 35 and born in 1951 Sweden in a non-stable population:
  
```{r, fig.height=10, fig.width=12}
swe_coh$kins_by_age_ego %>% 
  filter(cohort == 1951, age_ego == 35) %>% 
  select(kin, total) %>% 
  plot_diagram()
```

Finally, we show the expected number of ascendants and descendants by Ego's birth cohort in any given year:

```{r}
swe_coh$kins_by_age_ego %>%
  rename_kin() %>% 
  mutate(cohort = factor(cohort)) %>% 
  ggplot(aes(year,total,color=cohort)) +
  scale_y_continuous(labels = seq(0,3,.2),breaks = seq(0,3,.2))+
  geom_line(size=1)+
  geom_vline(xintercept = 2019, linetype=2)+
  facet_wrap(~kin,scales = "free")+
  labs(x = "Year", "Kin count") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

## Period perspective

For a **period perspective**, we re-run the `kins()` function, this time with a vector of `ego_year` values:

```{r, out.width = '100%'}
system.time(
swe_period <- 
  kins(
    U = U
    , f = f
    , pi = pi
    , stable = F
    , birth_female = 1
    # Note that we use the 'ego_year' parametersa as input
    , ego_year = c(1891,1921,1951,2010,2050,2080,2120)
    # We're only interested in certain kin ties
, selected_kins = c("d","gd","ggd","m","gm","ggm","os","ys","oa","ya")
)
)

```

Now, we plot the expected values by Ego's age at different time-points:

```{r}
swe_period$kins_by_age_ego %>%
  rename_kin() %>% 
  mutate(year = factor(year)) %>% 
  ggplot(aes(age_ego,total,color=year)) +
  geom_line(size=1)+
  facet_wrap(~kin, scales = "free")+
  labs(x = "Ego's age", "Kin count") + 
  theme_bw() +
  theme(legend.position = "bottom")
```

# In the pipeline

1. Implement multi-stage [@caswell_formal_2020] and two-sex models [@caswell2021formal_two-sex]
1. Improve Keyfitz kinship diagrams
1. Create a *Human Kinship Database*?

<!-- 1. Functions to transform common demographic data sources (HMD, HFD, UNWPP) to DemoKin input -->
  
# Get involved!
  
`DemoKin` is giving its first steps.
Please contact us via email, or create an issue or submit a pull request on GitHub if you have any suggestion: 
  
- https://github.com/IvanWilli/DemoKin
- ivanwilliams1985[at]gmail.com
- alburezgutierrez[at]demogr.mpg.de

# Citation

Williams, Iván, Alburez-Gutierrez, Diego and Xi Song. (2021) "DemoKin: An R package to compute kinship networks in stable and non-stable populations." URL: <https://github.com/IvanWilli/DemoKin>.

## References